Donkeys to Monoids: A Computational Framework for Dynamic Semantics

### 2013-05-09 paper.tex
Annotated appendix

Since we are going to experiment with converting language into logic, a natural place to start is with classical logic. Classical logic, also called Predicate logic (PL), is the logic of conjunction, disjunction and negation extended with universal and existential quantifiers.

In the introduction chapter we saw that PL was rather lacking in terms of compositionality and we predicted that what we really wanted was an `update logic' (a logic with update semantics) of some sort. Update semantics are well known and studied in computer language theory, and have in the past couple of decades found their way into a wider range of philosophies.

The idea of all of those is a `growth of information in time'. Starting from little, and with each statement adding more. Information is here seen in the information theoretic way, where no information is when everything is possible ($\top$) and total information is when nothing is possible ($\bot$).

@@

\subsection{The problem with disjunction}

@@

\subsection{Connection with first order logic}

@@

\subsection{Monadic variants}

Let's call them `Visser variants' or `VDPL'. Hm, I like that.

What should we call those? Visser calls them DPL variants, e.g. DPL$_{\Bowtie\triangle}$

@@

\subsection{Montague semantic}
Montague semantics is a theory of natural language that believes its semantics can be described thoroughly in terms of standard mathematical models. Logician Richard Montague (1930-1971) famously wrote
\begin{quotation}
There is in my opinion no important theoretical difference between natural languages and the artificial languages of logicians; indeed I consider it possible to comprehend the syntax and semantics of both kinds of languages with a single natural and mathematically precise theory. \cite{montague1970universal}
\end{quotation}

@@

\subsection{Weak and strong readings}

@@

\subsection{Dependency grammars}

@@

\tikzstyle{level 1}=[level distance=2.5cm, sibling distance=3.5cm]
\begin{figure}
\centering
\begin{tikzpicture}[grow=down, sloped]
\node {root}
    child {
        node {owns}
        child {
            node {If}
            edge from parent
            node[above] {mark}
        }
        child {
            node {farmer}
            child {
                node {a}
                edge from parent
                node[above] {det}
            }
            edge from parent
            node[above] {nsubj}
        }
        child {
            node {beats}
            child {
                node {he}
                edge from parent
                node[above] {nsubj}
            }
            child {
                node {it}
                edge from parent
                node[above] {dobj}
            }
            edge from parent
            node[above] {advcl}
        }
        child {
            node {donkey}
            child {
                node {a}
                edge from parent
                node[above] {det}
            }
            edge from parent
            node[above] {dobj}
        }
        edge from parent
        node[above] {root}
    };
\end{tikzpicture}
\caption{M1} \label{fig:M1}
\end{figure}

@@

\newpage
\section{Appendix}
\subsection{Derivation of DPL preconditions}
\begin{align}
\alpha\models\langle\exists x\rangle\phi
 & \leftrightarrow \exists\beta (\alpha[\exists x]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\forall\omega (\omega \neq x \rightarrow \alpha_{\omega} = \beta_{\omega}) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\exists\nu (\forall\omega (\alpha[x:=\nu]_{\omega} = \beta_{\omega})) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\exists\nu (\alpha[x:=\nu] = \beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\nu (\exists\beta (\alpha[x:=\nu] = \beta \wedge \beta\models\phi)) \nonumber\\
 & \leftrightarrow \exists\nu (\alpha[x:=\nu]\models\phi) \nonumber\\
 & \leftrightarrow \alpha\models\exists x (\phi) \\
%
\intertext{Here we take $\alpha[x:=\nu]$ to mean the $\alpha$ with $\alpha_x$ assigned to the value of $\nu$.}
%
\alpha\models\langle\psi_1\cdot\psi_2\rangle\phi
 & \leftrightarrow \exists\beta (\alpha[\psi_1\cdot\psi_2]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta(\exists\gamma (\alpha[\psi_1]\gamma \wedge \gamma[\psi_2]\beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\gamma (\alpha[\psi_1]\gamma \wedge \exists\beta(\gamma[\psi_2]\beta \wedge \beta\models\phi)) \nonumber\\
 & \leftrightarrow \exists\gamma (\alpha[\psi_1]\gamma \wedge \gamma\models\langle\psi_2\rangle\phi) \nonumber\\
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\langle\psi_2\rangle\phi
%
\intertext{c}
%
\alpha\models\langle\psi_1 \cup \psi_2\rangle\phi
 & \leftrightarrow \exists\beta (\alpha[\psi_1 \cup \psi_2]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta ((\alpha[\psi_1]\beta \vee \alpha[\psi_2]\beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha[\psi_1]\beta \wedge \beta\models\phi) \vee \exists\beta (\alpha[\psi_2]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow (\alpha\models\langle\psi_1\rangle\phi) \vee (\alpha\models\langle\psi_2\rangle\phi) \nonumber\\
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi
%
\intertext{d}
%
\alpha\models\langle P(x_1,\dots,x_n)\rangle\phi
 & \leftrightarrow \exists\beta (\alpha[P(x_1,\dots,x_n)]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha = \beta \wedge P(\alpha_{x_1},\dots,\alpha_{x_n}) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow P(\alpha_{x_1},\dots,\alpha_{x_n}) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\models P(x_1,\dots,x_n) \wedge \phi
%
\intertext{f}
%
\alpha\models\langle\neg\psi\rangle\phi
 & \leftrightarrow \exists\beta (\alpha[\neg\psi]\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha=\beta \wedge \neg\exists\gamma(\alpha[\psi]\gamma) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \neg\exists\gamma(\alpha[\psi]\gamma) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \neg\exists\gamma(\alpha[\psi]\gamma \wedge \gamma\models\top) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\not\models\langle\psi\rangle\top \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\models \neg\langle\psi\rangle\top \wedge \phi
%
\intertext{e}\nonumber
\end{align}

>>>>

In these chapters I'm going to describe the logical theories by Visser and earlier dynamic semantists. I will introduce some addtions of my own and prove interesting properties about them. Finally, but most importantly, I will go into depth with the tools and ideas we need to write a strong logification framework.

In the previous chapter I showed how First Order Logic lack certain compositional properties. A possible solution to this problem is to use a so called `update logic' (a logic with update semantics). Update semantics are logics inspired on the semantics of programming languages where a variable, once introduced, will stick around for a period not know at the time of it's introduction.

The idea of those logics is a `growth of information in time'. Starting from no information, and with each statement adding more. Information is here seen in the information theoretic way, where no information is when everything is possible ($\top$) and total information is when nothing is possible ($\bot$).

@@

\subsubsection{The problem with disjunction}

@@

\subsubsection{Connection with first order logic}

@@

\subsection{Donkey Monads*}

@@

\subsection{Montague Semantics}
Montague semantics is a theory of natural language that believes its semantics can be described thoroughly in terms of standard mathematical models. Logician Richard Montague (1930-1971) famously wrote
\begin{quotation}
There is in my opinion no important theoretical difference between natural languages and the artificial languages of logicians; indeed I consider it possible to comprehend the syntax and semantics of both kinds of languages with a single natural and mathematically precise theory. \cite{montague1970universal}
\end{quotation}

@@

\subsubsection{Weak and strong readings*}

@@

\subsection{Dependency grammars*}

@@

\tikzstyle{level 1}=[level distance=2cm, sibling distance=3.5cm]
\tikzstyle{level 2}=[level distance=2.5cm, sibling distance=3.5cm]
\tikzstyle{level 3}=[level distance=2cm, sibling distance=3.5cm]
\begin{figure}
\centering
\begin{tikzpicture}[grow=down, sloped]
\node {root}
    child {
        node {owns}
        child {
            node {If}
            edge from parent
            node[above] {mark}
        }
        child {
            node {farmer}
            child {
                node {a}
                edge from parent
                node[above] {det}
            }
            edge from parent
            node[above] {nsubj}
        }
        child {
            node {beats}
            child {
                node {he}
                edge from parent
                node[above] {nsubj}
            }
            child {
                node {it}
                edge from parent
                node[above] {dobj}
            }
            edge from parent
            node[above] {advcl}
        }
        child {
            node {donkey}
            child {
                node {a}
                edge from parent
                node[above] {det}
            }
            edge from parent
            node[above] {dobj}
        }
        edge from parent
        node[above] {root}
    };
\end{tikzpicture}
\caption{M1} \label{fig:M1}
\end{figure}

@@

\newpage
\section{Appendix}
\subsection{Derivation of DPL preconditions}
Remember that we have defined, for assignments $\alpha$ and $\beta$, DPL formula $\psi$ and FOL formula $\phi$: $\alpha\models\langle\psi\rangle\phi$ to mean $\exists\beta(\alpha\llbracket\psi\rrbracket\beta\wedge\beta\models\phi)$. $\alpha\models\phi$ is just the usual $\models$ from FOL.

Since the proof is over the syntax of DPL, we use $\llbracket\cdot\rrbracket$ as the interpretation function so that $\alpha\llbracket\psi\rrbracket\beta$ is the FOL statement `$(\alpha,\beta)\in\Psi$', where $\Psi$ is the relation defined by $\psi$.
\begin{align}
%
\intertext{Let's start out with the most tricky derivation. How does the resetting, unbounded $\exists$ of DPL translate into the defining, bounded $\eixsts$ of FOL?}
%
\alpha\models\langle\exists x\rangle\phi
 & \leftrightarrow \exists\beta (\alpha\llbracket\exists x\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\forall\omega (\omega \neq x \rightarrow \alpha_{\omega} = \beta_{\omega}) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\exists\nu (\forall\omega (\alpha[x:=\nu]_{\omega} = \beta_{\omega})) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\exists\nu (\alpha[x:=\nu] = \beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\nu (\exists\beta (\alpha[x:=\nu] = \beta \wedge \beta\models\phi)) \nonumber\\
 & \leftrightarrow \exists\nu (\alpha[x:=\nu]\models\phi) \nonumber\\
 & \leftrightarrow \alpha\models\exists x (\phi) \\
%
\intertext{Where we have taken $\alpha[x:=\nu]$ to mean the $\alpha$ with $\alpha_x$ assigned to the value of $\nu$.
The derivation for $\cdot$ goes more smoothly:}
%
\alpha\models\langle\psi_1\cdot\psi_2\rangle\phi
 & \leftrightarrow \exists\beta (\alpha\llbracket\psi_1\cdot\psi_2\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta(\exists\gamma (\alpha\llbracket\psi_1\rrbracket\gamma \wedge \gamma\llbracket\psi_2\rrbracket\beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\gamma (\alpha\llbracket\psi_1\rrbracket\gamma \wedge \exists\beta(\gamma\llbracket\psi_2\rrbracket\beta \wedge \beta\models\phi)) \nonumber\\
 & \leftrightarrow \exists\gamma (\alpha\llbracket\psi_1\rrbracket\gamma \wedge \gamma\models\langle\psi_2\rangle\phi) \nonumber\\
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\langle\psi_2\rangle\phi
%
\intertext{$\cup$ interestingly doesn't translate to $\langle\psi_1\vee\psi_2\rangle\phi$ as we might have expected:}
%
\alpha\models\langle\psi_1 \cup \psi_2\rangle\phi
 & \leftrightarrow \exists\beta (\alpha\llbracket\psi_1 \cup \psi_2\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta ((\alpha\llbracket\psi_1\rrbracket\beta \vee \alpha\llbracket\psi_2\rrbracket\beta) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha\llbracket\psi_1\rrbracket\beta \wedge \beta\models\phi) \vee \exists\beta (\alpha\llbracket\psi_2\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow (\alpha\models\langle\psi_1\rangle\phi) \vee (\alpha\models\langle\psi_2\rangle\phi) \nonumber\\
 & \leftrightarrow \alpha\models\langle\psi_1\rangle\phi \vee \langle\psi_2\rangle\phi
%
\intertext{For predicates it's mostly a matter of juggling variables, values and assigments:}
%
\alpha\models\langle P(x_1,\dots,x_n)\rangle\phi
 & \leftrightarrow \exists\beta (\alpha\llbracket P(x_1,\dots,x_n)\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha = \beta \wedge P(\alpha_{x_1},\dots,\alpha_{x_n}) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow P(\alpha_{x_1},\dots,\alpha_{x_n}) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\models P(x_1,\dots,x_n) \wedge \phi
%
\intertext{$\neg$ is interesting because it has to create a scope:}
%
\alpha\models\langle\neg\psi\rangle\phi
 & \leftrightarrow \exists\beta (\alpha\llbracket\neg\psi\rrbracket\beta \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \exists\beta (\alpha=\beta \wedge \neg\exists\gamma(\alpha\llbracket\psi\rrbracket\gamma) \wedge \beta\models\phi) \nonumber\\
 & \leftrightarrow \neg\exists\gamma(\alpha\llbracket\psi\rrbracket\gamma) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \neg\exists\gamma(\alpha\llbracket\psi\rrbracket\gamma \wedge \gamma\models\top) \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\not\models\langle\psi\rangle\top \wedge \alpha\models\phi \nonumber\\
 & \leftrightarrow \alpha\models \neg\langle\psi\rangle\top \wedge \phi
%
\intertext{Because $\psi$ is this way `evaluated on $\top$ only', we know that no variables introduced in $\psi$ can affect $\phi$.}\nonumber
\end{align}
